---
title: "AE & Motivation to Stop BZRAs"
author: "PO Couture"
format: html
editor: visual
---

## AEs & Motivation to Stop BZRAs

This is the code I will use for the publication about people 65+ and their reported adverse effects and their motivation to stop using or continue using BZRAs to see if any of these factors are important. Previous research has pointed to factors associated with LT use or success in cessation depending on the intervention but no study has been found simply describing if the experience of AEs and someones motivation to stop using BZRAs after LT use are associated.

## Loading The Data

Loading int the SIMOA_Report dataset.

```{r}
#| label: Loading The Data

library(readr)

SIMOA <- read_csv("SIMOA Report.csv")
View(SIMOA)

```

## Eligible Participants

```{r}
#| label: Inclusion/Exclusion Criteria
######
# This section of the code will ensure participants are 65+ and have taken a BZRA for sleep rather than another medication. I also show how many people I lose at each stage of the process to be able to report these later and make the code more ytransparent especially if it needs to be provided ot reviewers/ the journal.
######

library(tidyverse)  

# Original count
n_original <- nrow(SIMOA)

# After age filtering
SIMOA_age_filtered <- SIMOA %>%
  filter(age_cat == 1 | (age_cat == 0 & age >= 65))
n_after_age <- nrow(SIMOA_age_filtered)

# After both filters
SIMOA <- SIMOA_age_filtered %>%
  filter(rowSums(select(., starts_with("c_sp___"))[, 1:14] == 1, na.rm = TRUE) > 0)
n_final <- nrow(SIMOA)

# Report results
cat("Original sample:", n_original, "\n")
cat("After age filtering:", n_after_age, "(removed:", n_original - n_after_age, ")\n")
cat("After c_sp filtering:", n_final, "(removed:", n_after_age - n_final, ")\n")
cat("Total removed:", n_original - n_final, "\n")

```


## 


